\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage[round,authoryear]{natbib}
\usepackage[hidelinks]{hyperref}

\title{Axodus: A Prototype-Stage Conceptual Architecture for Traceability and Accountability in Federated Modular Digital Institutions}
\author{Mauricio Z. Filho\\
\small Independent Researcher, Axodus\\
\small Correspondence: mzfshark@gmail.com\\
\small Project discussion: Telegram @thinkincoin\\
\small Author handle: @mzfshark}
\date{Draft v0.4}

\begin{document}
\maketitle

\begin{center}
\fbox{\parbox{0.88\linewidth}{\centering
\textbf{Status:} Ready for author review of internal v0.4 draft; not ready for peer review, external circulation, or publication.\\
\textbf{Document type:} Prototype-stage conceptual architecture and focused research-agenda paper.\\
\textbf{License:} CC BY 4.0.}}
\end{center}

\begin{abstract}
Human-AI-assisted digital institutions become difficult to govern when
constitutional constraints, bounded local governance domains, modular
capabilities, and review obligations must be coordinated within the same
decision environment. Existing literatures on socio-technical systems,
decentralized governance, provenance, accountable algorithms, contestability,
and human-AI interaction clarify parts of this problem, but they do not by
themselves provide a compact conceptual architecture for traceability,
accountability, contestability, and reconstructability across
constitutionally constrained modular institutions. This paper presents Axodus
as a prototype-stage conceptual architecture for that setting. Its central
contribution is a six-layer governance-oriented model relating a
constitutional center, bounded local governance domains, modular
institutional capabilities, evidence and provenance, bounded AI assistance,
and risk or contestability controls. Participation readiness and internal
maturity distinctions are treated as supporting mechanisms rather than
co-equal contributions. Axodus is described only as a conceptual,
prototype-stage model; the paper reports no deployment, empirical validation,
observed institutional use, or operational readiness.
\end{abstract}

\noindent\textbf{Keywords:} socio-technical systems; federated digital
institutions; human-AI coordination; traceability; accountability;
governance; contestability

\section{Introduction}

Digital institutions increasingly combine distributed authority, modular
capabilities, public records, and automated assistance. Governance becomes
harder when these elements must remain reviewable across a constitutional
center and bounded local governance domains rather than inside a single
authority chain. In that setting, the institutional challenge is not only
whether a decision can be documented, but whether later reviewers can
reconstruct which authority applied, which bounded capability was invoked,
what evidence was considered, how AI assistance shaped the record, and how
correction, challenge, or appeal could proceed. Socio-technical systems
research treats those questions as jointly organizational and technical rather
than reducible to one domain alone \citep{baxter2011sociotechnical}.

Relevant literatures address parts of this problem without yet yielding a
single architecture for accountable governance across constitutionally
constrained modular ecosystems. Research on decentralized governance shows
that self-executing rules do not remove interpretation, amendment, or social
coordination \citep{hassan2021dao,hsieh2018bitcoin}. Human-AI interaction
guidance emphasizes review, correction, and expectation setting rather than an
undifferentiated transfer of authority \citep{amershi2019guidelines}.
Provenance and credentialing standards help distinguish evidence lineage and
machine verification from institutional judgment
\citep{w3c2013provdm,w3c2025vcdm}. Yet those literatures do not by themselves
provide a compact architecture for traceability, accountability,
contestability, and reconstructability in constitutionally constrained,
modular, human-AI-assisted institutions.

This paper addresses that gap by introducing Axodus as a prototype-stage
conceptual architecture for human-AI-assisted digital institutions under
bounded governance constraints. The project originates from a multi-year
independent research and design effort. The present contribution remains
conceptual: it does not report a platform in real institutional use, a
functioning institutional environment, evaluation results, or independently
tested controls.

The contribution is architecture-level design reasoning rather than an
implementation claim. The paper proposes a layered model for how a
constitutional center, bounded local governance domains, modular
institutional capabilities, evidence packages, attributable AI assistance,
and review paths may be arranged into one inspectable decision structure.
Participation readiness, internal maturity distinctions, and risk boundaries
remain supporting components that frame how this structure might later be
evaluated, narrowed, or rejected.

Implementation, deployment, institutional use, and empirical validation are
outside the paper's scope. The manuscript also excludes token design,
financial operations, educational-distribution mechanics, security
guarantees, and legal conclusions.

\section{Background and Related Work}

\subsection{Socio-technical systems and federated digital institutions}

A socio-technical perspective asks how roles, incentives, work practices,
information, and technology jointly shape outcomes. Baxter and Sommerville
argue for integrating organizational concerns with systems engineering rather
than treating them as a late adoption problem
\citep{baxter2011sociotechnical}. Axodus adopts this perspective by modeling
governance and accountability as architectural elements, not external policy
documents attached after implementation.

The term \emph{federated modular digital institution} is used here for an
organized system in which a central constitutional layer defines shared
constraints, while bounded local governance domains may configure modular
institutional capabilities. This use is conceptual. It does not imply a
deployed federation, legal federal status, or operating multi-tenant
platform.

\subsection{Decentralized governance, modular coupling, and review}

Hassan and De Filippi describe a DAO as a blockchain-based system in which
people coordinate through decentralized governance and self-executing rules
\citep{hassan2021dao}. Hsieh et al.\ distinguish machine consensus from the
social consensus required to alter protocols and organizational arrangements
\citep{hsieh2018bitcoin}. These distinctions caution against equating
technical distribution with accountable governance.

Smart-contract-based financial markets provide one domain-specific example in
which composability can propagate dependencies across connected protocols
\citep{schar2021defi}. Systems-theoretic safety research instead offers the
broader claim needed here: failures in complex socio-technical systems can
arise from interactions, control mismatches, and poorly governed coupling
rather than from isolated component defects alone
\citep{leveson2012engineering}. The present paper therefore treats that
domain only as one motivating example of dependency propagation, not as the
paper's institutional center.

\subsection{Human-AI oversight and accountable decision support}

Guidelines for human--AI interaction emphasize communicating system
capabilities, supporting correction, and learning from behavior over time
\citep{amershi2019guidelines}. At the organizational level, the NIST AI Risk
Management Framework treats governance, mapping, measurement, and management
as connected risk functions \citep{tabassi2023airmf}. These sources motivate
an architecture in which AI outputs are attributable inputs to decisions
remain attributable, reviewable inputs to human decisions rather than
sovereign decisions in themselves.

Constitutional AI uses written principles to guide model behavior through
supervised learning and AI feedback \citep{bai2022constitutional}. Axodus uses
\emph{constitutional} differently: it refers to institutional rules about
authority, prohibitions, amendments, review, and appeal. The proposed model
does not claim to train or align an AI model through Constitutional AI.

\subsection{Provenance, credentials, and institutional judgment}

The W3C PROV Data Model describes provenance through entities, activities,
agents, derivations, attribution, and responsibility relationships
\citep{w3c2013provdm}. This vocabulary is useful for distinguishing source
artifacts, transformations, automated assistance, and responsible roles in a
decision record. It supports reconstruction of how an output was produced,
but does not by itself establish that the decision was legitimate,
contestable, or accountable.

The W3C Verifiable Credentials Data Model separates the cryptographic
verification of a credential from validation of whether its claims are fit for
a verifier's purpose \citep{w3c2025vcdm}. This matters here because machine-
verifiable eligibility artifacts would not by themselves establish fair
assessment, legitimate authorization, or accountable participation
boundaries.

Work on accountable algorithms likewise argues that automated decision
processes must be assessable against external legal or policy standards and
that transparency alone is insufficient \citep{kroll2017accountable}.
Provenance is therefore one input to accountability, not its substitute.

\subsection{Contestability and reviewable governance}

Contestability-by-design literature identifies safeguards, human review and
intervention, and third-party scrutiny as design features for challenge and
correction \citep{alfrink2023contestable}. This framing is directly relevant
to federated modular governance because a traceable record remains
insufficient if a person affected by a decision cannot identify the
applicable rule, challenge evidence, or reach an appropriately independent
reviewer.

\section{Problem Statement}

The architecture addresses a design problem rather than an established
empirical effect:

\begin{quote}
How can a constitutionally constrained federated modular digital institution
preserve traceability, accountability, contestability, and
reconstructability when bounded local governance domains configure modular
institutional capabilities through human-AI-assisted decision processes?
\end{quote}

Five tensions structure the problem.

\begin{enumerate}[leftmargin=*]
\item \textbf{Central constraints versus local autonomy.} A constitutional
layer may preserve coherence, while bounded local governance domains still
require meaningful discretion over local decisions and configurations.
\item \textbf{Modular capabilities versus governability.} Service separation
may limit direct coupling, while modular selection can create hidden
dependency chains, ambiguous authority, and fragmented review paths.
\item \textbf{Automation versus accountable disposition.} AI assistance can
reduce analytical load, but responsibility can become unclear when its
outputs influence authorization, escalation, or review.
\item \textbf{Participation versus bounded readiness.} Open participation may
support legitimacy, while consequential governance actions may require
reviewable preparation, accommodations, and appeal.
\item \textbf{Documentation versus institutional reality.} A detailed
specification can improve inspection while also creating a misleading
impression of implementation completeness or legitimate authority.
\end{enumerate}

Axodus is proposed as one architecture for making these tensions inspectable
and reviewable. It is not presented as a complete or empirically tested
solution.

\section{Conceptual Architecture}

The model comprises six logical layers. They are presented as conceptual
design surfaces for allocating authority, structuring records, and preserving
reviewability, not as an implementation topology.

Information and control move through the architecture in different ways.
Control begins with constitutional constraints that define what local
decisions, capability selections, and escalation paths are in scope.
Evidence packages, provenance records, advisory analysis, rationale, and
review records then make those decisions reconstructable for later review.
Risk boundaries constrain permissible state transitions, while contestability
paths preserve challenge, correction, escalation, and appeal.

\noindent\textbf{Illustrative synthetic scenario - bounded-governance
exception review.} A mock bounded local governance domain requests a
temporary exception affecting one modular institutional capability. The
request includes the applicable constitutional rule, local authority basis,
declared conflicts, and a controlled evidence package. A simulated AI review
flags one conflict but omits a dependency between the requested
configuration and a review requirement. The designated human reviewer
withholds disposition, records the omission, and escalates the matter because
a risk boundary is triggered. An independent appeal role can then reconstruct
which constitutional constraint applied, what local choice was proposed, what
the simulated AI review changed, and why escalation occurred. This scenario
is an analytical illustration, not evidence of an Axodus implementation or
institutional process.

\begin{enumerate}[leftmargin=*]
\item constitutional governance defines the applicable roles, decision
rights, prohibitions, amendment rules, and appeal conditions;
\item bounded local governance domains interpret that scope for a local
decision context and may propose a configuration or exception within
constitutional limits;
\item modular institutional capabilities express what is being selected,
reviewed, or changed;
\item the evidence and provenance layer links source materials, authority
statements, conflicts, transformations, and responsible roles;
\item human-AI assistance may analyze that material, but only as attributable
advisory input to human disposition; and
\item risk-boundary and contestability controls constrain state transitions
and keep challenge, correction, escalation, and appeal available.
\end{enumerate}

\subsection{Constitutional governance layer}

The constitutional governance layer is a conceptual constraint surface that
defines higher-order rules: roles, decision rights, prohibitions, amendment
procedures, escalation paths, and appeal conditions. Its purpose is to make
non-delegable institutional constraints visible before local configuration
begins and to frame how later review can distinguish a permitted action from
an out-of-scope one.

In this proposal, the layer structures accountability boundaries rather than
serving as a deployed constitution, legal instrument, active authority, or
enforcement body. A reviewable action therefore depends on both institutional
justification and whatever technical authorization a future implementation
might use; neither alone would establish legitimacy.

\subsection{Tenant governance layer}

The Tenant governance layer represents a bounded local governance domain
operating under constitutional constraints. A Tenant may configure local
processes, select service nuclei, and allocate decision tasks only within the
scope permitted by the constitutional layer. The model treats Tenants as a
conceptual governance construct rather than as evidence of deployed multi-
tenant infrastructure, real communities, or operating customer entities.

\subsection{Service nuclei layer}

Service nuclei are abstract functional modules through which institutional
capabilities can be selected, bounded, and governed. The paper does not
present a product catalog. Instead, it treats a service nucleus as a
conceptual unit that should declare its purpose, inputs, outputs, delegated
authority, evidence requirements, dependencies, escalation conditions, and
failure behavior.

A minimal functional taxonomy includes the following component types:

\begin{itemize}[leftmargin=*]
\item an \textbf{authority and role registry} for decision rights and
delegations;
\item an \textbf{evidence registry} and \textbf{AI-assistance provenance
layer} for sources, transformations, and attributable automated inputs;
\item a \textbf{decision-rationale record} and \textbf{conflict and disclosure
register} for the basis and conditions of a decision;
\item a \textbf{review and appeal path} and \textbf{risk-boundary controller}
for challenge, escalation, and bounded action; and
\item \textbf{participation-readiness support} for educational preparation
before selected consequential activities.
\end{itemize}

These are component types rather than claims about a definitive Axodus module
inventory or implementation state.

\subsection{Evidence and provenance layer}

The evidence and provenance layer links proposals, supporting materials,
authority statements, declared conflicts, AI-generated analyses, decisions,
and subsequent reviews. Its design objective is reconstructability: a
reviewer should be able to ask what was known, who was authorized, what
configuration was proposed, what assistance was used, and why an outcome
followed.

At the conceptual level, source documents and outputs can be treated as
entities, analysis and review steps as activities, and human or automated
participants as agents. Derivation and attribution relations can then connect
an output to its inputs and responsible roles \citep{w3c2013provdm}. These
concepts specify what a reviewable record should express so that a later
reviewer can reconstruct the path from source material to disposition. They
do not by themselves prove that a record is true, complete, tamper-resistant,
or accountable in the stronger normative sense.

Accordingly, the layer supports later inspection rather than automatic
validation. It does not prescribe a storage system, cryptographic proof,
implemented evidence infrastructure, or integrity guarantee, and it does not
convert record linkage into a full judgment about whether the decision was
substantively justified.

\subsection{AI-assistance layer}

AI assistance remains advisory. It may support bounded tasks such as document
comparison, proposal summarization, inconsistency detection, constitutional
analysis support, adversarial checking, and planning-oriented review support.
Each use should declare the task, source material, output, uncertainty where
available, human reviewer, and disposition so that later reviewers can
inspect how advisory analysis entered the record. Humans retain decision
authority and responsibility.

This layer is intentionally role-abstract. The paper does not use named
agent labels, does not claim autonomous execution authority, and does not
claim that AI assistance itself enforces governance, issues binding
judgments, or replaces human institutional review.

\subsection{Risk-boundary and contestability layer}

The risk-boundary and contestability layer constrains permissible actions and
keeps review paths available after a decision is proposed or made. A boundary
should identify the responsible role, triggering condition, required
evidence, permitted exception, and escalation path. Contestability requires a
participant or reviewer to identify the applicable rule, challenge evidence,
correct records, disclose conflicts, and request review by an appropriately
independent role.

Within this proposal, reviewer independence means separation from the
original decision, disclosure of conflicts, and recusal when a reviewer
participated in the contested analysis. Escalation paths should identify
where a review moves when authority, expertise, or capacity is insufficient.
Human oversight becomes nominal when a reviewer lacks authority, time, access
to evidence, or a practical ability to change the outcome
\citep{tabassi2023airmf,amershi2019guidelines}. Contestability-by-design
literature identifies human review, intervention, and third-party scrutiny
as relevant design features \citep{alfrink2023contestable}.

\section{Participation Readiness as a Supporting Mechanism}

A participation-readiness mechanism is proposed only as a supporting
institutional interface for learning, assessment, bounded eligibility, and
appeal in selected consequential contexts. It is not the paper's central
contribution and must not be treated as a general social ranking system.

The W3C credential model could inform portable representations, but
verification of an issuer and credential does not validate educational
quality or justify authorization \citep{w3c2025vcdm}. The governance problem
therefore includes who defines the criteria, how accommodations are provided,
how a participant can contest an outcome, and whether the requirement is
proportionate to the decision context.

The central hypothesis is that educational readiness may improve bounded
participation in selected high-consequence contexts. Competing hypotheses
remain equally important: the mechanism may exclude participants, centralize
curriculum power, encourage test gaming, or create privacy risks. It should
be rejected or redesigned if those harms cannot be controlled.

\section{Internal Maturity Distinction and Scope Reduction}

The Axodus internal governance process distinguishes two orthogonal maturity
dimensions. \textbf{L-Level} refers to organizational, documentation,
governance, and authority maturity. \textbf{D-Level} refers to development,
implementation, technical validation, and operational maturity. L-Level does
not measure code, and D-Level does not measure governance. Neither dimension
authorizes production or sensitive action; production status and action
authority require separate gates.

The taxonomy is relevant here only as a guard against category error. It
helps prevent documentation from being mistaken for implementation, local
implementation from being mistaken for production, or maturity from being
mistaken for authority. The paper does not assign maturity levels to
components and does not present the taxonomy as a validated measurement
instrument.

Scope reduction is therefore a design requirement rather than a late-stage
optimization. Initial prototypes should start with decisions for which
reconstruction and appeal are materially important, use a minimum sufficient
decision-and-evidence record, separate advisory AI outputs from authoritative
human dispositions, and remove components that do not improve
reconstructability or contestability.

\section{Prototype Status}

Axodus is presented in this paper as a conceptual, prototype-stage model. The
public repository presently supports editorial artifacts, governance records,
and research documentation. It does not supply evidence of an integrated
implementation, functioning federated platform, production operation, user
adoption, empirical performance, or effective controls.

The absence of those claims is substantive. Architectural consistency can be
reviewed before implementation, but it cannot substitute for observed
behavior. Future versions should maintain a component-by-component evidence
table distinguishing designed, prototyped, tested, deployed, and
independently evaluated states.

\section{Evaluation Plan}

The evaluation program is intentionally prioritized rather than broad.

Before either initial study, the initial empirical stage requires a controlled
environment rather than a deployed institutional system. The environment
should combine synthetic Tenant-governance scenarios, controlled evidence
packages, simulated AI-assistance records, planted errors, and a high-fidelity
mock-up or wizard-of-oz review interface. A preregistered protocol should
define reviewer tasks and proposed measurement criteria. These criteria are
study designs, not reported results.

\paragraph{Primary initial study---RQ1: Traceability and reconstruction.}
Does the proposed evidence model help independent reviewers reconstruct
Tenant-level service-nucleus governance decisions more accurately and
efficiently than a document-only baseline? The controlled environment would
support matched governance scenarios and blinded review. Proposed measures are
reconstruction completeness, authority-conflict detection, missing-evidence
detection, time to reconstruct rationale, provenance error detection,
reviewer disagreement, and correct classification of whether an appeal can
proceed. This is the paper's primary evaluation priority because it most
directly tests the central claim about traceability and accountability in
federated modular governance.

\paragraph{Secondary initial study---RQ3: Human-AI reliance in advisory
review.} For bounded proposal-analysis tasks, how do provenance and mandatory
human disposition affect error detection, calibrated reliance, override
behavior, and review quality? The simulated AI records should include planted
errors, ambiguous cases, and contested recommendations. Proposed measures are
planted-error detection, unsupported-output detection, override or acceptance
behavior by condition, rationale quality, reviewer disagreement, and
correction after provenance information is shown. This is a secondary
priority because the architecture depends on bounded AI assistance being
reviewable rather than merely available.

\paragraph{Staged future evaluation.} Evaluation should progress from
controlled artifact inspection to bounded reviewer studies using synthetic
scenarios, then to controlled prototype evaluation under a separately approved
protocol. Any later institutionally situated study would require additional
governance, ethics, legal, privacy, and evidence gates. Participation
readiness, maturity reliability, and broader institutional-governance
questions remain later research topics.

\section{Limitations and Threats to Validity}

This paper is conceptual and still broader than a mature external article
would ideally be. The architecture can appear coherent because it has not yet
encountered implementation constraints, institutional conflict, workload
pressure, or field conditions that would test whether its proposed review
paths are workable across constitutional and Tenant-level decisions.

The public evidence base for Axodus is currently limited to design
documentation. There are no results with which to assess feasibility,
usability, governance legitimacy, calibrated reliance, contestable appeals,
or social impact. The paper reports no deployment validation, no
institutional validation, no production operation, no token-design or token-
distribution claims, no operational-finance claims, no security guarantees,
and no legal or regulatory conclusion.

Human oversight can become nominal when reviewers lack time, authority,
evidence access, or practical power to change the outcome. AI assistance may
be unreliable, difficult to calibrate, or difficult to audit under realistic
workload and time pressure. Weak contestability or poorly designed appeal
paths could make traceability visible without making accountability
meaningful.

Participation-readiness mechanisms present particular risks of exclusion,
surveillance, curriculum capture, and false confidence. Modular separation
can also hide systemic coupling, while constitutional constraints can be
overly rigid or too ambiguous for effective local governance. The proposed
architecture may be too complex to govern in practice without substantial
scope reduction.

The synthetic scenario is explanatory rather than empirical. Provenance
vocabulary does not establish record integrity or accountability, and the
proposed measures do not demonstrate performance. Institutionally situated
evaluation may also expose conflicts, incentives, and workload effects that a
controlled artifact cannot reproduce.

The paper does not provide financial, security, investment, or legal advice.
Its terminology may not map cleanly onto every jurisdiction or research
community.

\section{Research Agenda}

Near-term work should prioritize a minimal decision-and-evidence model,
synthetic Tenant-governance scenarios, and bounded AI-assistance tasks
suitable for initial traceability and reliance studies. Additional literature
hardening around governance quality, reviewer independence, empirical
institutional appeals, and participation-readiness validity remains a later
requirement before any broader external review is considered.

\section{Conclusion}

Axodus is proposed as a prototype-stage conceptual architecture for making
human-AI-assisted governance more traceable and accountable in federated
modular digital institutions. Its contribution is not a product catalog or
operational claim, but an inspectable arrangement linking constitutional
constraints, bounded Tenant autonomy, service nuclei, evidence and
provenance, advisory AI assistance, and review paths.

The next research step is to instantiate the minimum decision-and-evidence
model in a controlled experimental environment and test whether independent
reviewers can reconstruct and challenge Tenant-level governance decisions
under the proposed record structure.

\noindent References are maintained in \texttt{references.bib}. All entries
cited in this draft have corresponding internal verification records in
\texttt{../../research/bibliography-verified.md}; final metadata rechecking is
required before any submission, public release, or external circulation.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
